Stochastic Variance-Reduced Policy Gradient

### 2018-02-09 svrpg.tex
baseline

The main limitation of plain policy gradient is the high variance of these estimators.
The na\"ive approach of increasing the batch size is not an option in RL due to the high cost of collecting samples, \ie by interacting with the environment.
For this reason, literature has focused on the introduction of baselines (\ie functions $b : \mathcal{S} \times \mathcal{A} \to \realspace$) aiming to reduce the variance~\citep[\eg][]{williams1992simple,Peters2008reinf,Thomas2017actionbaseline,wu2018variance}, see Appendix~\ref{A:gradient_estimators} for a formal definition of $b$.
{\color{red}
These baselines are usually designed in order to minimize the variance of the gradient estimate but even them need to be estimated from data, partially reducing they effectiveness.
On the other hand, there has been a surge of recent interest in variance reduction techniques for gradient optimization in supervised learning.
Although these techniques have been mainly derived for finite-sum problems, we will show in Section~\ref{sec:alg} how they can be used in RL.
In particular, we will shown that the proposed SVRPG algorithm can take the best of both worlds (\ie SL and RL) since it can be plugged over a policy gradient estimate using baselines.
The next section has the aim to describe variance reduction techniques for finite-sum problems. In particular, we will present the SVRG algorithm that is at the core of this work.
}

@@

\section{Conclusion}
In this paper, we introduced SVRPG, a variant of SVRG designed explicitly for RL problems.
The control problem considered in the paper has a series of difficulties that are not common in SL.
Among them, non-concavity and approximate estimates of the FG have been analysed independently in SL~\citep[\eg][]{allen2016variance,reddi2016stochastic,harikandeh2015stopwasting} but never combined.
Nevertheless, the main issue in RL is the non-stationarity of the sampling process since the distribution underlying the objective function is policy-dependent.
We have shown that by exploiting importance weighting techniques, it is possible to overcome this issue and preserve the unbiasedness of the corrected gradient.
We have additionally shown that, under mild assumptions that are often verified in RL applications, it is possible to derive convergence guarantees for SVRPG.
%{\color{red}SVRG has been previously used in RL for policy evaluation~\citep{du2017svrgpe} and as alternative optimization method for TRPO~\citep{xu2017svrgtrpo}.
%The first case is much more similar to the SL case since the problem is stationary and may be made concave, removing several issues we have faced here.
%In the second case, it was used as a better optimization technique to solve the per-iteration problem imposed by TRPO.
%}
%\todomatout{Repetition of Section 6?}
Finally, we have empirically shown that practical variants of the theoretical SVRPG version can outperform classical actor-only approaches on benchmark tasks.
Preliminary results support the effectiveness of SVRPG also with a commonly used baseline for the policy gradient.
Despite that, we believe that it will be possible to derive a baseline specifically designed for SVRPG in order to jointly exploit the RL structure and the SVRG idea.
\todomat{Other future work}

@@

\section{Practical SVRPG versions}\label{app:practicalsvrpg}
\begin{algorithm}[h]
        \begin{algorithmic}
        \STATE \textbf{Input:} A gradient estimate $g_t$ and parameters $\beta_1$, $\beta_2$, $\epsilon$ and $\alpha$.
        \STATE $m_t = \beta_1 m_{t-1} + (1 - \beta_1) g_t$
        \STATE $v_t = \beta_2 v_{t-1} + (1 - \beta_2) g_t \circ g_t$ ($\circ$ is the Hadamard (component-wise) product)
        \STATE $\hat{m}_t = \dfrac{m_t}{1 - \beta^t_1}$
        \STATE $\hat{v}_t = \dfrac{v_t}{1 - \beta^t_2}$
        \STATE $\Delta(g_t) = \dfrac{\alpha}{\sqrt{\hat{v}_t} + \epsilon} \hat{m}_t$
        \STATE \textbf{Return:} The increment $\Delta(g_t)$  of the parameters.
\end{algorithmic}
 \caption{
 \label{A:adam}
 Adam}
\end{algorithm}
\todopir{Explain how we use adam and when/if it is reset on snapshots}

\section{Experimental Details}\label{app:exp}
Used environments:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : 4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
	\item \emph{Mujoco Swimmer}: 13-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
	\item \emph{Mujoco Half Cheetah}: 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}

>>>>

The main limitation of plain policy gradient is the high variance of these estimators.
The na\"ive approach of increasing the batch size is not an option in RL due to the high cost of collecting samples, \ie by interacting with the environment.
For this reason, literature has focused on the introduction of baselines (\ie functions $b : \mathcal{S} \times \mathcal{A} \to \realspace$) aiming to reduce the variance~\citep[\eg][]{williams1992simple,Peters2008reinf,Thomas2017actionbaseline,wu2018variance}.
{\color{red}
This baselines are usually designed in order to minimize the variance of the gradient estimate but even them need to be estimated from data, partially reducing they effectiveness.
On the other hand, there has been a surge of recent interest in variance reduction techniques for gradient optimization in supervised learning.
Although these techniques have been mainly derived for finite-sum problems, we will show in Section~\ref{sec:alg} how they can be used in RL.
In particular, we will shown that the proposed SVRPG algorithm can take the best of both worlds (\ie SL and RL) since it can be plugged over a policy gradient estimate using baselines.
The next section has the aim to describe variance reduction techniques for finite-sum problems. In particular, we will present the SVRG algorithm that is at the core of this work.
}

@@

\section{Conclusion}
In this paper, we introduced SVRPG, a variant of SVRG designed explicitly for RL problems.
The control problem considered in the paper has a series of difficulties that are not common in SL.
Among them, non-concavity and approximate estimates of the FG have been analysed independently in SL~\citep[\eg][]{allen2016variance,reddi2016stochastic,harikandeh2015stopwasting} but never combined.
Nevertheless, the main issue in RL is the non-stationarity of the sampling process since the distribution underlying the objective function is policy-dependent.
We have shown that by exploiting importance weighting techniques, it is possible to overcome this issue and preserve the unbiasedness of the corrected gradient.
We have additionally shown that, under mild assumptions that are often verified in RL applications, it is possible to derive convergence guarantees for SVRPG.
{\color{red}SVRG has been previously used in RL for policy evaluation~\citep{du2017svrgpe} and as alternative optimization method for TRPO~\citep{xu2017svrgtrpo}.
The first case is much more similar to the SL case since the problem is stationary and may be made concave, removing several issues we have faced here.
In the second case, it was used as a better optimization technique to solve the per-iteration problem imposed by TRPO.
}
\todomatout{Repetition of Section 6?}
Finally, we have empirically shown that practical variants of the theoretical SVRPG version can outperform classical actor-only approaches on benchmark tasks.
\todomat{add future work}

@@

\section{Practical SVRPG versions}\label{app:practicalsvrpg}
\begin{algorithm}[h]
        \begin{algorithmic}
        \STATE \textbf{Input:} A gradient estimate $g_t$ and parameters $\beta_1$, $\beta_2$, $\epsilon$ and $\alpha$.
        \STATE $m_t = \beta_1 m_{t-1} + (1 - \beta_1) g_t$
        \STATE $v_t = \beta_2 v_{t-1} + (1 - \beta_2) g_t \circ g_t$ ($\circ$ is the Hadamard (component-wise) product)
        \STATE $\hat{m}_t = \dfrac{m_t}{1 - \beta^t_1}$
        \STATE $\hat{v}_t = \dfrac{v_t}{1 - \beta^t_2}$
        \STATE $\Delta(g_t) = \dfrac{\alpha}{\sqrt{\hat{v}_t} + \epsilon} \hat{m}_t$
        \STATE \textbf{Return:} The increment $\Delta(g_t)$  of the parameters.
\end{algorithmic}
 \caption{
 \label{A:adam}
 Adam}
\end{algorithm}
\subsection{Learning Rates}
Let us give a deeper insight on the two different learning rate used by our algorithm. We have already said that we use two different ADAM one related to the snapshots and another one related to the sub-iterations. In this way tha ADAM associated to the snapshots will take into account only the snapshots history through the parameters $m_{t-1}$ and $v_{t-1}$. The same reasoning holds for the sub-iterations. This, of course, allows our algorithm to take into account the different variabilities in a more accurate manner.
\subsection{Baseline}
For all batch gradient-based algorithms, we use the same time-varying feature encoding for the linear baseline:
\begin{align*}
\psi_{s,t}=concat(s, s \odot s, 0.01t, (0.01t)^2, (0.01t)^3,1)
\end{align*}
where $s$ is the state vector and $\odot$ is the element-wise product. In the case of our algorithm the model representing the baseline is fitted every snapshot so we have:\\
\begin{align*}
        \nabla J(\pi^T) =\frac{1}{N} \sum_{n=1}^{N}
                \sum_{h=0}^H \left(\sum_{k=0}^h \nabla \log\pi^T(y_k^n)\right) (\gamma^h \Reward(y_h^n)-b_{sn})
          + \frac{1}{B} \sum_{n=1}^{B}
                \sum_{h=0}^H \left(\sum_{k=0}^h \nabla \log\pi^T(z_k^n)\right) (\gamma^h \Reward(z_h^n) - b_{si})\\ - \frac{1}{B} \sum_{n=1}^{B}
                \sum_{h=0}^H \left(\sum_{k=0}^h \nabla \log\pi^T(z_k^n)\right) (\gamma^h \Reward(z_h^n) - b_{si})\omega(z_{0:h}|\pi^B,\pi^T)
\end{align*}
Where $b_{sn}$ is the prediction given by the baseline at the snapshot, whereas $b_{si}$ is the prediction given by the baseline at a given subiteration. 
\section{Experimental Details}\label{app:exp}
Used environments:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : 4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
	\item \emph{Mujoco Swimmer}: 13-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
	\item \emph{Mujoco Half Cheetah}: 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}
